Computing likelihoods under the SMC for a general class of ARGs.

### 2025-02-14 paper.tex
Pass through the intro

\section{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now intense interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Numerous applications taking advantage of this detailed history
in inferred ARGs are now appearing~\citep{
stern_approximate_2019,
osmond_estimating_2021,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
ignatieva_ongoing_2022,
wang_complex_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
huang_estimating_2024,
grundler_geographic_2024,
korfmann_simultaneous_2024,
deraje_inferring_2024}
and it seems likely that many more will
follow~\citep{harris_database_2019,harris_using_2023}.
Although there are many different approaches to ARG
inference~\citep{wong_general_2023}, two broad classes of method
have been the focus of most recent interest.

@@

The second class of inference method that has been of recent interest 
is based on heuristics and approximate models.
Relate~\citep{speidel_method_2019},
tsinfer~\citep{kelleher_inferring_2019} and 
ARG-Needle~\citep{zhang_biobank-scale_2023} work on quite different 
principles, but share some common properties. Firstly, they are all 
hueristic and approximation driven, favouring computational
scalability over statistical rigor, and lacking a direct 
connection to population genetic theory. 
Secondly, they all produce a single
ARG as output, inferring a deterministic point estimate with no means 
of quantifying uncertainty. Thirdly (and most importantly for the 
purposes of this paper) the ARGs they produce lack substantial
detail about ancestral recombination events: the estimates are
rough sketches, hopefully capturing large-scale properties, but lacking
in the finer detail of outputs of ARGweaver or SINGER.

@@

[Needs more work]
Despite these significant advantages over hueristic methods, however,
hueristic methods remain of interest because of the profound difference
in computational scalability between these approaches. Even with the 
major advances made in this direction by SINGER, ARG inference is 
only feasible for hundreds of samples. Tsinfer and ARG-Needle have 
been applied to datasets of hundreds of \emph{thousands} of samples.
With the explosive growth in dataset
sizes in recent years~\citep[e.g.][]{bycroft_genome_2018,halldorsson_sequences_2022,
all_genomic_2024} there is a pressing need for more statistical
rigor in large scale ARG inference. 
% This inability to connect the results of large-scale inference methods
% with a generative model
% is a fundamental limitation, and limits both the utility of 
% current point estimates as well as the possibilities for further
% inference. Specifically, we are unable to quantify uncertainty 
% around heuristic point estimates. We cannot explore ARG space
% around the point estimates using (e.g. MCMC), as we have no way definition
% of improvement. We cannot compare models of demography, or evaluate the 
% goodness of fit for model estimates.
A key problem facing the field is that the ARGs estimated by heuristic large-scale 
methods cannot be connected to population genetics theory because 
we cannot compute their likelihood under the SMC. The classical 
method~\citep{kuhner_maximum_2000} assumes complete knowledge about
recombination events, and there is no clear way to compute a likelihood for 
the approximate structures output by hueristic methods.

Computing the likelihood of a putative ARG under a model is a fundamental element of
probabilistic inference. The classical Kuhner, Yamato \& Felsenstein
\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
works by assigning a probablility to each recombination and common
ancestor event, and to the inter-event  waiting times. The likelihood of an 
ARG is then a product over all these three classes of term.
In essence, computing the KYF likelihood requires
``replaying'' the  state of Hudson's simulation
algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
compute the exact configuration of lineages and distribution
of ancestral material at each time point~\citep{baumdicker_efficient_2021}.
It therefore requires complete
detail about each of these events to be stored. 
Other approaches to defining the likelihood of an ARG under a coalescent-like 
model follow the same basic event-by-event 
strategy~\citep[e.g.][]{guo_recombination-aware_2022}.
[Are there others?  The bacterial stuff from Didelot etc seem to focus on
a tree-by-tree likelihood, rather than global graph, so not clear it's the 
same thing]
Computing the likelihood under the SMC rather than the Hudson coalescent 
simplifies bookkeeping somewhat, 
but a complete event-by-event description of the 
input ARG is still required.

Although the KYF formulation is a natural and elegant way to describe 
the likelihood of an ARG, such complete detail is not present 
in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
Simulators such as the classical \texttt{ms}
program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
which contain no information about recombination events (other than their
breakpoints) and where internal nodes marking coalesces are unlabelled.
The \texttt{msprime}
simulator~\citep{kelleher_efficient_2016,baumdicker_efficient_2021} 
by default outputs a more complete and efficient representation of 
the simulated ancestry, but still does not provide the exhaustive
detail required.
Indeed, providing the information required to 
support likelihood calculations originally motivated the addition of
the \texttt{record\_full\_arg}
option to \texttt{msprime}~\citep{baumdicker_efficient_2021}.
Forwards-time simulations that output 
ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} similarly do not
contain the level of detail required to compute a likelihood under
the KYF approach.

[Things to include: locally unary nodes rather than explicit events.
edge-centric formulation vs event centric. Graceful degradation of
performance with incomplete ARGs vs complete inability to compute.]

In this paper we provide a necessary first step towards this goal
of statistically rigorous large-scale ARG inference, by defining a function
to compute the likelihood of a general class of ARG under the SMC.
To do this, we rederive the SMC under a backwards-time formulation,
modelling the process in terms of lineages and [something] rather
than events. Using this novel formulation, we then show how to 
easily compute the likelihood of any ARG under the SMC and outline 
an efficient algorithm to do so. We illustrate the flexibility of this
approach via [some examples] and conclude by discussing avenues
for future research opened up.

>>>>

\section{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now intense interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024,nielsen_inference_2025}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Numerous applications taking advantage of this detailed history
in inferred ARGs are now appearing~\citep{
stern_approximate_2019,
osmond_estimating_2021,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
ignatieva_ongoing_2022,
wang_complex_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
huang_estimating_2024,
grundler_geographic_2024,
korfmann_simultaneous_2024,
deraje_inferring_2024,
speidel2025high}
and it seems likely that many more will
follow~\citep{harris_database_2019,harris_using_2023}.
Although there are many different approaches to ARG
inference~\citep{wong_general_2023}, two broad classes of method
have been the focus of most recent interest.

@@

The second class of inference method that has been of recent interest 
is based on heuristics and approximate models.
Relate~\citep{speidel_method_2019},
tsinfer~\citep{kelleher_inferring_2019},
ARG-Needle~\citep{zhang_biobank-scale_2023} 
and Threads~\citep{gunnarsson_scalable_2024}
work on quite different 
principles, but share some common properties. 
Firstly, they are all 
hueristic and approximation driven, favouring computational
scalability over statistical rigor.
Secondly, they all produce a single
ARG as output, inferring a deterministic point estimate with no means 
of quantifying uncertainty. Thirdly (and most importantly for the 
purposes of this paper) the ARGs they produce lack substantial
detail about ancestral recombination events: the estimates are
rough sketches, hopefully capturing large-scale properties, but lacking
in the finer detail of outputs of ARGweaver or SINGER.

@@

Despite these significant advantages of rigorous inference, however,
hueristic methods remain of interest because of the profound difference
in computational scalability. Although SINGER is a major step forward
in terms of scalability over ARGweaver and other methods, inference is still 
only feasible for hundreds of samples. 
Tsinfer, ARG-Needle and Threads have 
been applied to datasets of hundreds of \emph{thousands} of samples.
With the explosive growth in dataset
sizes in recent years~\citep{caulfield2017national,turnbull2018100,
bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole,
all2024genomic,stark2024call,cook2025our}
there is a pressing need for more statistical
rigor in large scale ARG inference.
% This inability to connect the results of large-scale inference methods
% with a generative model
% is a fundamental limitation, and limits both the utility of 
% current point estimates as well as the possibilities for further
% inference. Specifically, we are unable to quantify uncertainty 
% around heuristic point estimates. We cannot explore ARG space
% around the point estimates using (e.g. MCMC), as we have no way definition
% of improvement. We cannot compare models of demography, or evaluate the 
% goodness of fit for model estimates.
A key problem facing the field is that the ARGs estimated by heuristic large-scale 
methods lack a concrete connection to population genetic theory because
we cannot compute their likelihood under the SMC or similar models. 
% The classical method~\citep{kuhner_maximum_2000} assumes complete knowledge about
% recombination events, and there is no clear way to compute a likelihood for 
% the approximate structures output by hueristic methods.

Computing the likelihood of an ARG under a model is a fundamental element of
probabilistic inference. The classical Kuhner, Yamato \& Felsenstein
\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
works by assigning a probablility to each recombination and common
ancestor event, and to the inter-event  waiting times. The likelihood of an 
ARG is then a product over all these three classes of term.
In essence, computing the KYF likelihood requires
``replaying'' the  state of Hudson's simulation
algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
compute the exact configuration of lineages and distribution
of ancestral material at each time point~\citep{baumdicker_efficient_2021}.
It therefore requires complete
detail about each of these events to be stored. 
Other approaches to defining the likelihood of an ARG under a coalescent-like 
model follow the same basic event-by-event 
strategy~\citep[e.g.][]{guo_recombination-aware_2022}.
% [Are there others?  The bacterial stuff from Didelot etc seem to focus on
% a tree-by-tree likelihood, rather than global graph, so not clear it's the 
% same thing]
Computing the likelihood under the SMC rather than the Hudson coalescent 
simplifies bookkeeping somewhat,  % Dunno what I'm talking about here
but a complete event-by-event description of the 
input ARG is still required.

Although the KYF formulation is a natural and elegant way to describe 
the likelihood of an ARG, such complete detail is not present 
in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
Simulators such as the classical \texttt{ms}
program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
which does not contain sufficient detail for likelihood computation.
The \texttt{msprime}
simulator~\citep{kelleher_efficient_2016,baumdicker_efficient_2021} 
by default outputs a more complete and efficient representation of 
the simulated ancestry, but still does not provide the exhaustive
detail required.
Indeed, providing the information required to 
support likelihood calculations originally motivated the addition of
the \texttt{record\_full\_arg}
option to \texttt{msprime}~\citep{baumdicker_efficient_2021}.
Forwards-time simulations that output 
ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} similarly do not
typically contain the level of detail required to compute a likelihood under
the KYF approach.

% [Things to include: locally unary nodes rather than explicit events.
% edge-centric formulation vs event centric. Graceful degradation of
% performance with incomplete ARGs vs complete inability to compute.]

In this paper we provide a new formulation of ARG likelihood which 
has much more lenient input requirements than KYF, and gracefully handles
incomplete and imprecise ARGs.
% In this paper we provide a necessary first step towards this goal
% of statistically rigorous large-scale ARG inference, by defining a function
% to compute the likelihood of a general class of ARG under the SMC.
To do this, we rederive the SMC under a backwards-time formulation,
modelling the process in terms of lineages and [something] rather
than events. Using this novel formulation, we then show how to 
easily compute the likelihood of any ARG under the SMC and outline 
an efficient algorithm to do so. We illustrate the flexibility of this
approach via [some examples] and conclude by discussing avenues
for future research opened up.
